Martingale representations for Markov chains with application to MCMC \tnoteref{mytitlenote}

### 2020-04-15 Article/elsarticle-mcmc-var.tex

\documentclass[bj]{imsart}
\makeatletter
\def\set@curr@file#1{%
  \begingroup
    \escapechar\m@ne
    \xdef\@curr@file{\expandafter\string\csname #1\endcsname}%
  \endgroup
}
\def\quote@name#1{"\quote@@name#1\@gobble""}
\def\quote@@name#1"{#1\quote@@name}
\def\unquote@name#1{\quote@@name#1\@gobble"}
\makeatother
\usepackage{graphics}

\usepackage{lineno,hyperref}
\usepackage{amsfonts,amssymb}
\usepackage{mathtools}
%\usepackage{mathbbol}
\usepackage{bbm}
\usepackage{multirow,algorithm2e}%
\usepackage{xargs}
%\usepackage{showlabels}

@@

\def\rmd{\mathrm{d}}
\def\rme{\mathrm{e}}
\def\rset{\mathbb{R}}
\def\NtrainPath{T}
\def\TrainSet{\mathcal{T}}
\modulolinenumbers[5]
\newcommand{\proofendsign}{$\Box$}
\newtheorem{thm}{Theorem}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{prop}[thm]{Proposition}
\newtheorem{remark}{Remark}
\newtheorem{cor}{Corollary}
\newtheorem{example}{Example}
\newenvironment{proof}{{\noindent \bf Proof }}
 {{\hspace*{\fill}\proofendsign\par\bigskip}}
\def\TV{\operatorname{TV}}
\newcommand{\jac}[1]{\operatorname{J} _{#1}}

@@

\begin{frontmatter}
\title{Variance reduction for MCMC methods via martingale representations}
\runtitle{Variance reduction for MCMC}
%\thankstext{T1}{This research was }

\begin{aug}
\author{\snm{D. Belomestny}\thanksref{a,b}\ead[label=e1]{denis.belomestny@uni-due.de}}
\and
\author{\snm{E. Moulines}\thanksref{c, b}\ead[label=e2]{eric.moulines@polytechnique.edu}}
\and
\author{\snm{N. Shagadatov}\thanksref{b}\ead[label=e3]{nshagadatov@edu.hse.ru}}
\and
\author{\snm{M. Urusov}\thanksref{a}\ead[label=e4]{murusov@uni-due.de}}

\runauthor{D. Belomestny et al}

\affiliation{Duisburg-Essen University, National Research University Higher School of Economics,  Ecole Polytechnique  and Duisburg-Essen University}

\address[a]{
Duisburg-Essen University,
Faculty of Mathematics,
D-45127 Essen
Germany }

\address[c]{
Centre de Math\'ematiques Appliqu\'ees, UMR 7641, Ecole Polytechnique,
France
}

\address[b]{
National University Higher School of Economics, Moscow, Russia
}

\end{aug}

\begin{abstract}
In this paper we propose an efficient variance reduction approach for  MCMC algorithms relying on a novel discrete time martingale representation for Markov chains. Our approach is fully non-asymptotic and does not require any type of ergodicity or special product structure of the underlying density.  By rigorously analyzing the  convergence properties of the proposed algorithm, we show that it's complexity is indeed asymptotically smaller than one of the original MCMC algorithm. The numerical performance of the new method is illustrated  in the case of Gaussian mixtures and Bayesian regression models.
\end{abstract}

\begin{keyword}[class=MSC]
\kwd[Primary ]{60G40}
\kwd{60G40}
\kwd[; secondary ]{91G80}
\end{keyword}

\begin{keyword}
\kwd{MCMC}
\kwd{variance reduction}
\kwd{martingale representation}
\end{keyword}

@@

@@

Monte Carlo integration typically has an error variance of the form
$\sigma^{2}/n,$ where $n$ is a sample size and \(\sigma^2\) is the variance of  integrand. We can make the variance smaller
by using a larger value of $n$.  Alternatively,
 we can  reduce $\sigma^2$ instead of increasing the sample size \(n\). To this end, one can try to construct
a new Monte Carlo experiment with the same expectation as the original one
but with a lower variance $\sigma^2$. Methods to achieve this are known as variance
reduction techniques. Variance reduction plays an important role in
Monte Carlo and Markov Chain Monte Carlo methods. Introduction to many of the variance reduction techniques can be found in \cite{christian1999monte}, \cite{rubinstein2016simulation}, \cite{GobetBook} and \cite{glasserman2013monte}. Recently one witnessed a revival of interest in efficient variance reduction methods  for MCMC algorithms, see for example \cite{dellaportas2012control}, \cite{mira2013zero}, \cite{brosse2018diffusion} and references therein.
\par
Suppose that we wish to compute $\pi(f):=\mathsf{E}\left[f(X)\right]$, where $X$
is a random vector-valued in $\mathcal{X}\subseteq\mathbb{R}^{d}$ with a density $\pi$ and $f:$
$\mathcal{X}\to\mathbb{R}$ with $f\in L^2(\pi)$.
The idea of the control variates variance reduction method
 is to find a cheaply computable random variable $\zeta$ with $\mathsf{E}[\zeta]=0$ and \(\mathsf{E}[\zeta^2]<\infty,\)
such that the variance of the r.v. $f(X)-\zeta$ is small.  The complexity of the problem of constructing classes $Z$ of control variates \(\zeta\) satisfying   $\mathsf{E}[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
For example, if \(\pi\) is analytically known and satisfies some regularity conditions, one can apply the well-known technique of  polynomial interpolation to construct control variates enjoying  some optimality properties, see, for example, Section~3.2 in \cite{dimov2008monte}. Alternatively, if an orthonormal system in \(L^2(\pi)\) is analytically available, one can build control variates \(\zeta\) as a linear combination of the corresponding basis functions, see \cite{GobetCV}. Furthermore, if \(\pi\) is known only up to a normalizing constant (which is often the case in Bayesian statistics), one can apply the recent approach of control variates  depending only on the gradient \(\nabla \log \pi\)  using Schr\(\ddot{\text{o}}\)dinger-type Hamiltonian operator in  \cite{mira2013zero} and Stein operator in \cite{brosse2018diffusion}.
In some situations \(\pi\) is not known analytically, but \(X\) can be represented as a function of  simple random variables with known distribution.
Such  situation arises, for example, in the case of functionals of  discretized diffusion processes. In this case a Wiener chaos-type decomposition can be used to construct control variates with nice theoretical properties, see \cite{belomestny2018stratified}.
Note that in order to compare different  variance reduction approaches, one has to analyze their complexity, that is, the number of numerical operations required to achieve a prescribed magnitude of the resulting variance.

The situation becomes much more difficult in the case of MCMC algorithms, where one  has to work with a
Markov chain \(X_p,\) \(p=0,1,2,\ldots,\) whose marginal distribution  converges  to \(\pi\) as time grows. One important class of the variance reduction methods in this case  is based on the Poisson equation for the corresponding  Markov chain. It was observed in~\cite{henderson1997variance}  that if a time-homogeneous Markov chain \((X_p)\) is stationary with stationary distribution \(\pi,\) then for any real-valued function \(G \in L^1(\pi) \) defined on the state space of the Markov chain \((X_p),\)  the function \(U(x) := G(x)-\mathsf{E}[G(X_{1})|X_0 = x]\) has zero mean with respect to \(\pi\).  The best choice for the function \(G\) corresponds to a solution of the Poisson equation  \(\mathsf{E}[G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\).  Moreover, it is also related to the minimal asymptotic variance in the corresponding central limit theorem, see \cite{duncan2016variance} and \cite{mira2013zero}.   Although the Poisson equation involves the quantity of interest \(\pi(f)\)  and can not be  solved explicitly in most cases, this idea still can be used to construct some  approximations for the optimal zero-variance control variates. For example,  \cite{henderson1997variance} proposed to compute approximations for the solution of the Poisson equation for specific Markov chains with particular emphasis on models arising in stochastic network theory. In \cite{dellaportas2012control} and \cite{brosse2018diffusion}  series-type control variates are introduced and studied for reversible Markov chains. It is assumed in \cite{dellaportas2012control}  that the one-step conditional expectations  can be computed explicitly  for a set of basis functions. The authors in \cite{brosse2018diffusion} proposed another approach tailored to diffusion setting which does not require the computation of integrals of basis functions and only involves  applications of the underlying generator.
\par
In this paper we focus on the  Langevin type algorithms which got much attention recently, see \cite{dalalyan2017theoretical,durmus:moulines:2017, MR2353037, MR3861816, MR2977986} and references therein. We propose  a generic variance reduction method for these and other types algorithms, which is purely non-asymptotic and does not require that    the conditional expectations of the corresponding Markov chain can be computed  or that the generator is known analytically. Moreover, we do not need to assume stationarity or/and sampling under the invariant distribution \(\pi.\) We rigorously analyse the convergence of the method and study its complexity. It is shown that  our variance-reduced Langevin algorithm outperforms the standard Langevin algorithms in terms of complexity.
\par
The paper is organized as follows.  In Section~\ref{sec:setup} we set up the problem and introduce some notations. Section~\ref{seq:mart_repr} contains a novel martingale representation and shows how this representation can be used for variance reduction. In Section~\ref{sec:ula_analysis} we analyze the performance of the proposed variance reduction algorithm in the case of Unadjusted Langevin Algorithm (ULA). Section~\ref{sec:coeff} studies the complexity of the variance reduced ULA. Finally, numerical examples are presented in Section~\ref{sec:num}.

\section{Setup}\label{sec:setup}
Let  \(\mathcal{X}\) be a domain in \( \mathbb{R}^d.\)  Our aim is to numerically compute  expectations of the form
\[
\pi(f)=\int_{\mathcal{X}} f(x)\pi(\rmd x),
\]
where \(f:\) \(\mathcal{X}\longrightarrow \mathbb{R}\) and \(\pi\) is a probability measure supported on \(\mathcal{X}.\)
If  the dimension of the space \(\mathcal{X}\) is large and \(\pi(f)\) can not be computed analytically, one can apply Monte Carlo methods. However, in many practical situations  direct sampling from \(\pi\) is impossible and this precludes the use of plain Monte Carlo methods in this case. One popular alternative to Monte Carlo  is Markov Chain Monte Carlo (MCMC), where one is looking for a discrete time  (possibly non-homogeneous) Markov chain   \((X_p)_{p\geq 0}\) such that \(\pi\) is its unique invariant measure. In this paper we study a class of MCMC algorithms with \((X_p)_{p\geq 0}\) satisfying the  the following recurrence relation:
\begin{equation}
\label{eq:chain_gen}
X_{p}=\Phi_{p}(X_{p-1},\xi_{p}),\quad p=1,2,\ldots ,\quad X_{0}=x_0,
\end{equation}
for some i.i.d.  random vectors \(\xi_p\in \mathbb{R}^m\) with distribution \(P_{\xi}\)
and some Borel-measurable
functions $\Phi_{p}\colon\mathcal{X}\times\mathbb{R}^{m}\to\mathcal{X}.$
In fact, this is quite general class of Markov chains (see Theorem~1.3.6 in ~\cite{moulines2018})
and many well-known MCMC algorithms can be represented in form \eqref{eq:chain_gen}.
Let us consider two popular examples.
\begin{example}[Unadjusted Langevin Algorithm]
\label{exam:langevin-algorithm}
Fix a sequence of positive time steps \((\gamma_p)_{p\geq 1}.\) Given a Borel function $\mu\colon\mathbb{R}^{d}\to\mathbb{R}^{d}$,
consider a non-homogeneous
discrete-time Markov chain $(X_{p})_{p\geq0}$ defined by
\begin{equation}\label{eq:chain}
X_{p+1}=X_{p}-\gamma_{p+1}\mu(X_{p})+\sqrt{\gamma_{p+1}}Z_{p+1},\end{equation}
where $\left(Z_{p}\right)_{p\geq1}$ is an i.i.d.\ sequence of $d$-dimensional
standard Gaussian random vectors. If
$\mu=(1/2) \nabla U$
for some continuously differentiable function $U,$ then Markov chain~\eqref{eq:chain} can be used to approximately sample from the density
\begin{equation}\label{eq:stationary_distr}
\pi(x)= \mathrm{const}\,\rme^{-U(x)}, \quad \mathrm{const}= \left. 1 \middle /{\int_{\mathbb{R}^{d}} \rme^{-U(x)}\, \rmd x,} \right.
\end{equation}
provided that \(\int_{\mathbb{R}^{d}} \rme^{-U(x)}\,\rmd x\) is finite. This method is usually referred to as
Unadjusted Langevin Algorithm (ULA).
In fact the Markov chain~\eqref{eq:chain}
arises as the Euler-Maruyama discretization
of the Langevin diffusion
\[
\rmd Y_t=-\mu(Y_t)\,\rmd t+ \rmd W_t
\]
with nonnegative time steps $(\gamma_p)_{p\ge1}$,
and,  under mild technical conditions, the latter Langevin diffusion admits $\pi$
of~\eqref{eq:stationary_distr}
as its unique invariant distribution; see \cite{dalalyan2017theoretical} and \cite{durmus:moulines:2017}.
\end{example}
\begin{example}[Metropolis-Adjusted Langevin Algorithm]
The Metropolis-Hastings algorithm
associated with a target density \(\pi\) requires the choice of a sequence of conditional densities  \((q_p)_{p\geq 1}\) also called proposal or candidate kernels. The transition from the value  of the Markov chain \(X_p\)  at time \(p\)
and its value at time \(p + 1\) proceeds via the following transition step:

\begin{algorithm}[H]
Given \(X_p=x\)\;
\begin{enumerate}
\item Generate \(Y_p\sim q_p(\cdot|x)\)\;
\item Put
X_{p+1}=
\begin{cases}
Y_p, & \texttt{ with probability } \alpha(x,Y_p),
x, &  \texttt{ with probability } 1-\alpha(x,Y_p),
\end{cases}
\alpha(x,y)=\min\left\{1,\frac{\pi(y)}{\pi(x)}\frac{q_p(x|y)}{q_p(y|x)}\right\}.
\end{enumerate}
\end{algorithm}
This transition is reversible with respect to \(\pi\) and therefore preserves the stationary density \(\pi\); see \cite[Chapter~2]{moulines2018}. If \(q_p\) have a wide enough
support to eventually reach any region
of the state space \(\mathcal{X}\) with positive mass
under \(\pi\), then this transition is irreducible and $\pi$ is a maximal irreducibility measure \cite{mengersen:tweedie:1996}. The  Metropolis-Adjusted Langevin algorithm (MALA) takes  \eqref{eq:chain} as proposal, that is,
q_p(y|x)=(\gamma_{p+1})^{-d/2}\boldsymbol{\varphi}\Bigl([y-x+\gamma_{p+1}\mu(x)]/\sqrt{\gamma_{p+1}}\Bigr),
$\boldsymbol{\varphi}(z)=(2\pi)^{-d/2} \exp\{-|z|^2/2\}$,
$z\in\mathbb R^d$, denotes the density of a $d$-dimensional
standard Gaussian random vector.  The MALA algorithms usually provide noticeable speed-ups in convergence for most problems. It is not difficult to see that the MALA chain can be compactly represented in the form
X_{p+1} &=X_p+\mathbbm{1}\bigl(U_{p+1}\leq \alpha(X_{p},Y_{p})\bigr)(Y_{p}-X_p),  \,
Y_{p}&=X_p-\gamma_{p+1}\mu(X_p)+\sqrt{\gamma_{p+1}}Z_{p+1},
where \((U_{p})_{p\geq 1}\) is an i.i.d. sequence of uniformly distributed on \([0,1]\) random variables independent of \((Z_p)_{p\geq 1}.\) Thus we recover \eqref{eq:chain_gen} with  \(\xi_p=(U_p,Z_p)\in \mathbb{R}^{d+1}\) and
\Phi_p(x,(u,z)^\top)=x+\mathbbm{1}\bigl(u\leq \alpha(x,x-\gamma_{p}\mu(x)+\sqrt{\gamma_{p}}z)\bigr)(-\gamma_{p}\mu(x)+\sqrt{\gamma_{p}}z).

@@

\begin{example}
  Let \((X_t)_{t\in [0,T]}\) be the unique strong solution to a  SDE of the form:
\label{eq:sde-inv}
dX_t=b(X_t)\, \rmd t+\sigma(X_t)\, \rmd W_t,\quad t\geq 0,	
where \(W\) is a standard \(\mathbb{R}^m\)-valued Brownian motion, \(b:\) \(\mathbb{R}^d\to \mathbb{R}^d\) and \(\sigma:\) \(\mathbb{R}^d\times \mathbb{R}^m \to \mathbb{R}^d\) are locally Lipschitz continuous functions with at most linear growth.
The process \((X_t)_{t\geq 0}\) is a Markov
process and let \(L\) denote its infinitesimal generator defined by
Lg=b^\top \nabla g+\frac{1}{2}\text{Tr}(\sigma^\top D^2g\sigma)
for any \(g\in C^2(\mathbb{R}^d).\)
If there exists a continuously twice differentiable Lyapunov function \(V:\) \(\mathbb{R}^d\to \mathbb{R}_{+}\) such that
\sup_{x\in \mathbb{R}^d} LV(x) <\infty,\quad \limsup_{|x|\to \infty} LV(x)<0,
then there is an invariant probability measure \(\pi\)  for \(X,\) that is, \(X_t\sim \pi\) for all \(t> 0\) if \(X_0\sim \pi.\) Invariant measures are crucial in the study of the long term behaviour of stochastic differential
systems \eqref{eq:sde-inv}.    Under some additional assumptions,  the invariant
measure \(\pi\) is  ergodic and this property
can be exploited  to compute  the integrals \(\pi(f)\) for \(f\in L^2(\pi)\) by means of ergodic averages. The idea is to replace the diffusion \(X\) by a (simulable) discretization scheme of the form (see e.g. \cite{MR3861816})
\bar X_{n+1}=\bar X_n+\gamma_{n+1} b(\bar X_n)+\sigma(\bar X_n)(W_{\Gamma_{n+1}}-W_{\Gamma_n}), \quad n\geq 0,\quad \bar X_0=X_0,
where \(\Gamma_n=\gamma_1+\ldots+\gamma_n\) and \((\gamma_n)_{n\geq 1}\) is a non-increasing sequence of time steps. Then for a function \(f\in L^2(\pi)\) we can approximate \(\pi(f)\) via
\pi_n^\gamma(f)=\frac{1}{\Gamma_n}\sum_{i=1}^n \gamma_{i}f(\bar X_{i-1}).
Due to typically high correlation between \(X_0,X_1,\ldots\)  variance reduction is of crucial importance here.
As a matter of fact, in many cases there is no explicit formula for the invariant measure and this makes the use of gradient  based  variance reduction techniques (see e.g. \cite{mira2013zero})   impossible in this case.
On the contrary, our method can be directly used to reduce the variance of the ergodic estimator \(\pi_n^\gamma\) without explicit knowledge of \(\pi.\)

>>>>

\documentclass[article]{elsarticle}
\makeatletter
\def\set@curr@file#1{%
  \begingroup
    \escapechar\m@ne
    \xdef\@curr@file{\expandafter\string\csname #1\endcsname}%
  \endgroup
}
\def\quote@name#1{"\quote@@name#1\@gobble""}
\def\quote@@name#1"{#1\quote@@name}
\def\unquote@name#1{\quote@@name#1\@gobble"}
\makeatother
\usepackage{graphics}
\usepackage{lineno,hyperref}
\usepackage{amsfonts,amssymb}
\usepackage{mathtools}
%\usepackage{mathbbol}
\usepackage{bbm}
\usepackage{multirow,algorithm2e}%
\usepackage{xargs}

@@

\def\bs{\mathbf{s}}
\def\TV{\operatorname{TV}}
\def\rmd{\mathrm{d}}
\def\rme{\mathrm{e}}
\def\rset{\mathbb{R}}
\def\nset{\mathbb{N}}
\def\NtrainPath{T}
\newcommandx{\TrainSet}[3][1=T,2=n,3=N]{\mathcal{T}_{#1,#2}^{#3}}
\newcommandx{\TrainAlg}[3][1=T,2=n,3=N]{\mathcal{\widetilde{T}}_{#1,#2}^{#3}}

\newcommandx{\EstFunc}[3][1=N]{\bar{Q}^{#1}_{#2,#3}}
\newcommandx{\EstFuncSmp}[3][1=N]{\widehat{Q}^{#1}_{#2,#3}}
\def\bk{\mathbf{k}}
\def\bX{\mathbf{X}}
\def\bY{\mathbf{Y}}
\def\Id{\operatorname{I}}
\newcommand{\jacob}[2]{\operatorname{J}_{#1}(#2)}
\newcommandx{\acoef}[4][1=N]{\bar{a}_{#2,#3}^{#1,#4}}
\newcommandx{\acoefest}[4][1=N]{\hat{a}_{#2,#3}^{#1,#4}}

\modulolinenumbers[5]

\journal{Journal of \LaTeX\ Templates}

%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%

%% Numbered
%\bibliographystyle{model1-num-names}

%% Numbered without titles
%\bibliographystyle{model1a-num-names}

%% Harvard
%\bibliographystyle{model2-names.bst}\biboptions{authoryear}

%% Vancouver numbered
%\usepackage{numcompress}\bibliographystyle{model3-num-names}

%% Vancouver name/year
%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}

%% APA style
%\bibliographystyle{model5-names}\biboptions{authoryear}

%% AMA style
%\usepackage{numcompress}\bibliographystyle{model6-num-names}

\newtheorem{thm}{Theorem}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{prop}[thm]{Proposition}
\newdefinition{remark}{Remark}
\newdefinition{cor}{Corollary}
\newdefinition{example}{Example}
\newproof{proof}{Proof}

%% `Elsevier LaTeX' style
\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{frontmatter}

\title{Martingale representations  for Markov chains   with application to MCMC \tnoteref{mytitlenote}}
\tnotetext[mytitlenote]{This work has been funded by the  Russian Academic Excellence Project '5-100'}

%% or include affiliations in footnotes:
\author[address1,address2]{D. Belomestny\corref{mycorrespondingauthor}}
\ead[url]{www.uni-due.de/~hm0124}
\ead{denis.belomestny@uni-due.de}

\author[address3,address2]{E. Moulines}
\author[address2]{N. Shagadatov}
\author[address1]{M. Urusov}
\cortext[mycorrespondingauthor]{Corresponding author}

\address[address1]{Duisburg-Essen University, Essen}
\address[address2]{National University Higher School of Economics, Moscow}
\address[address3]{Centre de Math\'ematiques Appliqu\'ees, UMR 7641, Ecole Polytechnique, France}

\begin{abstract}
In this paper we propose an efficient variance reduction approach for  MCMC algorithms relying on a novel discrete time martingale representation. Our approach is fully non-asymptotic and does not require any type of ergodicity or special product structure of the underlying density.   We rigorously analyze   complexity of the proposed algorithm and prove its reduction. The numerical performance is illustrated  in the case of Gaussian mixtures and binary regression.
\end{abstract}

\begin{keyword}
Monte Carlo\sep optimal stopping \sep regression \sep boosting
\MSC[2010] 65C05 \sep 60F05 \sep 62L10 \sep 65C40 \sep 60J05 \sep 93E35

\end{keyword}

@@

\linenumbers

@@

Monte Carlo integration typically has an error variance of the form
$\sigma^{2}/n,$ where $n$ is a sample size and \(\sigma^2\) is the variance of the integrand. We can make the variance smaller
by using a larger value of $n$.  Alternatively,
 we can  reduce $\sigma^2$ instead of increasing the sample size \(n\). To this end, one can try to construct
a new Monte Carlo experiment with the same expectation as the original one
but with a lower variance $\sigma^2$. Methods to achieve this are known as variance
reduction techniques. Variance reduction plays an important role in
Monte Carlo and Markov Chain Monte Carlo methods. Introduction to many of the variance reduction techniques can be found in \cite{christian1999monte}, \cite{rubinstein2016simulation} and \cite{glasserman2013monte}. Recently one witnessed a revival of interest in efficient variance reduction methods  for MCMC algorithms, see for example \cite{dellaportas2012control}, \cite{mira2013zero}, \cite{brosse2018diffusion} and references therein.

Suppose that we wish to compute $\pi(f):=\mathsf{E}\left[f(X)\right]$, where $X$
is a random vector-valued in $\mathcal{X}\subseteq\mathbb{R}^{d}$ with a density $\pi$ and $f:$
$\mathcal{X}\to\mathbb{R}$ with $f\in L^2(\pi)$.
The idea of the so-called control variates variance reduction method
 is to find a cheaply computable random variable $\xi$ with $\mathsf{E}[\xi]=0$ and \(\mathsf{E}[\xi^2]<\infty,\)
such that the variance of the r.v. $f(X)-\xi$ is small.  The complexity of the problem of constructing classes $\Xi$ of control variates \(\xi\) satisfying   $\mathsf{E}[\xi]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
For example, if \(\pi\) is analytically known and satisfies some regularity conditions, one can apply the well-known technique of  polynomial interpolation to construct control variates enjoying  some optimality properties, see, for example, Section~3.2 in \cite{dimov2008monte}. Alternatively, if an orthonormal system in \(L^2(\pi)\) is analytically available, one can build control variates \(\xi\) as a linear combination of the corresponding basis functions. Furthermore, if \(\pi\) is known only up to a normalizing constant (which is often the case in Bayesian statistics), one can apply the recent approach of
 \cite{mira2013zero} and further worked out in\cite{oates2017control}  suggesting  control variates  depending only on the gradient \(\nabla \log \pi.\)

In some situations \(\pi\) is not known analytically, but \(X\) can be represented as a function of  simple random variables with known distribution.
Such  situation arises, for example, in the case of functionals of  discretized diffusion processes. In this case a Wiener chaos-type decomposition can be used to construct control variates with nice theoretical properties, see \cite{belomestny2018stratified}.
Note that in order to compare different  variance reduction approaches, one has to analyze their complexity, that is, the number of numerical operations required to achieve a prescribed magnitude of the resulting variance.

The situation becomes much more difficult in the case of MCMC algorithms, where one  has to work with a
Markov chain \(X_p,\) \(p=0,1,2,\ldots,\) whose marginal distribution  converges  to \(\pi\) as time grows. One important class of the variance reduction methods in this case  is based on the so-called Poisson equation for the corresponding  Markov chain. It was observed in Henderson~\cite{henderson1997variance}  that if a time-homogeneous Markov chain \((X_p)\) is stationary with stationary distribution \(\pi,\) then for any real-valued function \(G \in L^1(\pi) \) defined on the state space of the Markov chain \((X_p),\)  the function \(U(x) := G(x)-\mathsf{E}[G(X_{1})|X_0 = x]\) has zero mean with respect to \(\pi\), provided that \(\pi(|G|) < \infty\).  The best choice for the function \(G\) corresponds to a solution of the so-called Poisson equation  \(\mathsf{E}[G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\).  In fact,  the Poisson equation leads to a zero-variance control variate for the empirical mean under \(\pi.\) Moreover, it is also related to the minimal asymptotic variance in the corresponding central limit theorem, see \cite{duncan2016variance} and \cite{mira2013zero}.   Although the Poisson equation involves the quantity of interest \(\pi(f)\)  and can not be  solved explicitly in most cases, this idea still can be used to construct some  approximations for the optimal zero-variance control variates. For example,  Henderson~\cite{henderson1997variance} proposed to compute approximations for the solution of the Poisson equation for specific Markov chains with particular emphasis on models arising in stochastic network theory. In \cite{dellaportas2012control} and \cite{brosse2018diffusion}  series-type control variates are introduced and studied for reversible Markov chains. It is assumed in \cite{dellaportas2012control}  that the one-step conditional expectations  can be computed explicitly  for a set of basis functions. The authors in \cite{brosse2018diffusion} proposed another approach tailored to diffusion setting which doesn't require the computation of integrals of basis functions and only involves  applications of the underlying generator.
\par
In this paper we focus on the  Langevin type algorithms which got much attention recently, see \cite{dalalyan2017theoretical,durmus:moulines:2017} and references therein. We propose  a generic variance reduction method for these and other types algorithms, which is purely non-asymptotic and does not require that    the conditional expectations of the corresponding Markov chain can be computed explicitely or that the generator is known analytically. Moreover, we do not need to assume stationarity or/and sampling under the invariant distribution \(\pi.\) We rigorously analyse the convergence of the method and study its complexity. It is shown that  our variance-reduced Langevin algorithm outperforms the standard Langevin algorithms in terms of complexity.
\par
The paper is organized as follows.  In Section~\ref{sec:setup} we set up the problem and introduce some notations. Section~\ref{seq:mart_repr} contains a novel martingale representation and shows how this representation can be used for variance reduction. In Section~\ref{sec:ula_analysis} we analyze the performance of the proposed variance reduction algorithm in the case of Unadjusted Langevin Algorithm (ULA). Section~\ref{sec:coeff} studies the complexity of the variance reduced ULA. Finally, numerical examples are presented in Section~\ref{sec:num}.
\section{Setup}\label{sec:setup}
Let  \(\mathcal{X}\) be a domain in \( \mathbb{R}^d.\)  Our aim is to numerically compute the expectations of the form
\pi(f)=\int_{\mathcal{X}} f(x)\pi(\rmd x),
where \(f:\) \(\mathcal{X}\longrightarrow \mathbb{R}\) and \(\pi\) is a probability measure supported on \(\mathcal{X}.\)
If  the dimension of the space \(\mathcal{X}\) is large and \(\pi(f)\) can not be computed analytically, one can apply Monte Carlo methods. However, in many practical situations  direct sampling from \(\pi\) is impossible and this precludes the use of plain Monte Carlo methods in this case. One popular alternative to Monte Carlo  is Markov Chain Monte Carlo, where one is looking for a discrete time  (possibly non-homogeneous) Markov chain   \((X_p)_{p\geq 0}\) such that \(\pi\) is its unique invariant measure. In this paper we study a class of MCMC algorithms with \((X_p)_{p\geq 0}\) satisfying the  the following recurrence relation:
\label{eq:chain_gen}
X_{p}=\Phi_{p}(X_{p-1},\xi_{p}),\quad p=1,2,\ldots ,\quad X_{0}=x_0,
for some i.i.d.  random vectors \(\xi_p\in \mathbb{R}^m\) with distribution \(P_{\xi}\)
and some Borel-measurable
functions $\Phi_{p}\colon\mathcal{X}\times\mathbb{R}^{m}\to\mathcal{X}.$
In fact, this is quite general class of Markov chains (see Theorem~1.3.6 in ~\cite{moulines2018})
and many well-known MCMC algorithms can be represented in form \eqref{eq:chain_gen}.
Let us consider two popular examples.
\begin{example}[Unadjusted Langevin Algorithm]
\label{exam:langevin-algorithm}
Fix a sequence of positive time steps \((\gamma_p)_{p\geq 1}.\) Given a Borel function $\mu\colon\mathbb{R}^{d}\to\mathbb{R}^{d}$,
consider a non-homogeneous
discrete-time Markov chain $(X_{p})_{p\geq0}$ defined by
\begin{equation}\label{eq:chain}
X_{p+1}=X_{p}-\gamma_{p+1}\mu(X_{p})+\sqrt{\gamma_{p+1}}Z_{p+1},\end{equation}
where $\left(Z_{p}\right)_{p\geq1}$ is an i.i.d.\ sequence of $d$-dimensional
standard Gaussian random vectors. If
${\mu=(1/2)\nabla U}$
for some continuously differentiable function $U,$ then Markov chain~\eqref{eq:chain} can be used to approximately sample from the density
\begin{equation}\label{eq:stationary_distr}
\pi(x)= \mathrm{const}\,\rme^{-U(x)}, \quad \mathrm{const}= \left. 1 \middle /{\int_{\mathbb{R}^{d}} \rme^{-U(x)}\, \rmd x,} \right.
provided that \(\int_{\mathbb{R}^{d}} \rme^{-U(x)}\,\rmd x\) is finite. This method is usually referred to as
Unadjusted Langevin Algorithm (ULA).
In fact the Markov chain~\eqref{eq:chain}
arises as the Euler-Maruyama discretization
of the Langevin diffusion
\rmd Y_t=-\mu(Y_t)\,\rmd t+ \rmd W_t
with nonnegative time steps $(\gamma_p)_{p\ge1}$,
and,  under mild technical conditions, the latter Langevin diffusion admits $\pi$
of~\eqref{eq:stationary_distr}
as its unique invariant distribution; see \cite{dalalyan2017theoretical} and \cite{durmus:moulines:2017}.

@@

\begin{example}[Metropolis-Adjusted Langevin Algorithm]
The Metropolis-Hastings algorithm
associated with a target density \(\pi\) requires the choice of a sequence of conditional densities  \((q_p)_{p\geq 1}\) also called proposal or candidate kernels. The transition from the value  of the Markov chain \(X_p\)  at time \(p\)
and its value at time \(p + 1\) proceeds via the following transition step:

\begin{algorithm}[H]
Given \(X_p=x\)\;
\begin{enumerate}
\item Generate \(Y_p\sim q_p(y|x)\)\;
\item Take
\begin{eqnarray*}
X_{p+1}=
\begin{cases}
Y_p, & \texttt{ with probability } \alpha(x,Y_p),
\\
x, &  \texttt{ with probability } 1-\alpha(x,Y_p),
\end{cases}
\end{eqnarray*}
where
\begin{eqnarray*}
\alpha(x,y)=\min\left\{\frac{\pi(y)}{\pi(x)}\frac{q_p(x|y)}{q_p(y|x)}\right\}.
\end{eqnarray*}
\end{enumerate}
\end{algorithm}
Then, as shown in Metropolis et al.~\cite{metropolis1953equation}, this transition is reversible with respect to \(\pi\) and therefore preserves the stationary density \(\pi\). If \(q\) has a wide enough
support to eventually reach any region
of the state space \(\mathcal{X}\) with positive mass
under \(\pi\), then this transition is irreducible and $\pi$ is a maximal irreducibility measure \cite{mengersen:tweedie:1996}. The  Metropolis-Adjusted Langevin algorithm (MALA) takes  \eqref{eq:chain} as proposal, that is,
\begin{eqnarray*}
q_p(y|x)=(\gamma_{p+1})^{-d/2}\boldsymbol{\varphi}\left([y-x+\gamma_{p+1}\mu(x)]/\sqrt{\gamma_{p+1}}\right),
\end{eqnarray*}
where
$\boldsymbol{\varphi}(z)=(2\pi)^{-d/2} \exp\{-|z|^2/2\}$,
$z\in\mathbb R^d$, denotes the density of a $d$-dimensional
standard Gaussian random vector.  MALA algorithms usually provide noticeable speed-ups in convergence for most problems. It is not difficult to see that the MALA chain can be compactly represented in the form
\begin{align*}
X_{p+1} &=X_p+\mathbbm{1}\bigl(U_{p+1}\leq \alpha(X_{p},Y_{p})\bigr)(Y_{p}-X_p),  \\
Y_{p}&=X_p-\gamma_{p+1}\mu(X_p)+\sqrt{\gamma_{p+1}}Z_{p+1},
\end{align*}
where \((U_{p})_{p\geq 1}\) is an i.i.d. sequence of uniformly distributed on \([0,1]\) random variables independent of \((Z_p)_{p\geq 1}.\) Thus we recover \eqref{eq:chain_gen} with  \(\xi_p=(U_p,Z_p)\in \mathbb{R}^{d+1}\) and
\begin{eqnarray*}
\Phi_p(x,(u,z)^\top)=x+\mathbbm{1}\bigl(u\leq \alpha(x,x-\gamma_{p}\mu(x)+\sqrt{\gamma_{p}}z)\bigr)(-\gamma_{p}\mu(x)+\sqrt{\gamma_{p}}z).
\end{eqnarray*}
\end{example}
